Numerical Stability of Detonations in White Dwarf Simulations

### 2019-02-23 papers/ignition/paper.tex
Remove tracked changes

Early studies on WD collisions \citep{rosswog:2009,raskin:2010,loren-aguilar:2010,
hawley:2012,garcia-senz:2013} typically had effective spatial resolutions in the
range 100--500 km \added{for the grid codes, and 10--100 km for the SPH codes}, and observed detonations that convert a large amount of
carbon/oxygen material into iron-group elements. These studies varied in methodology
(Lagrangian versus Eulerian evolution, nuclear network used) and did not closely agree
on the final result of the event (see Table 4 of \cite{garcia-senz:2013} for a summary).
\cite{kushnir:2013} argued that many of these simulations featured numerically unstable
evolution, ultimately caused by the zone size being significantly larger than the length
scale over which detonations form. The detonation length scale can vary widely based
on physical conditions \citep{seitenzahl:2009,garg:2017} but is generally \replaced{much
smaller than 100 km}{not larger than 10 km}.

In this paper, we attempt to find what simulation length scale is required to achieve
converged thermonuclear ignitions. \added{The inspiration for this work comes from our
simulations of WD collisions using the reactive hydrodynamics code \castro\
\citep{castro, astronum:2017}. We have done both 2D axisymmetric and 3D simulations
of collisions of $0.64\ \msolar$ carbon/oxygen WDs, and we were unable to achieved converged
simulations at any resolution we could afford to run (the best was an effective zone size of
0.25 km, using adaptive mesh refinement, for the 2D case). We were therefore forced to
turn to 1D simulations, where we can achieve much higher resolution (at the cost, of course,
of not being able to do a test that can be directly compared to multi-dimensional simulations).
Our goal is to record our thoughts on why we believe we did not achieve convergence.}

@@

Our test problem is inspired by \cite{kushnir:2013}, and very loosely approximates the
conditions of two $0.64\ \msolar$ WDs colliding head-on. The simulation domain is 1D with a
reflecting boundary at $x = 0$. For $x > 0$ there is a uniform fluid composed\footnote{
\added{We note that the WD simulations that inspired this were pure carbon/oxygen,
but for the setup we describe here, pure carbon/oxygen conditions do not detonate at all.
This is not particularly surprising, since the 1D setup is a very imperfect representation
of the real multi-dimensional case, and is missing multi-dimensional hydrodynamics that could
substantially alter the dynamical evolution. So we added a small amount of helium to ensure
that this setup ignites. Of course, there will likely be a small amount of helium present in
C/O white dwarfs as a remnant of the prior stellar evolution.}} (by mass)
of $50\%\, ^{12}$C, $45\%\, ^{16}$O, and $5\%\, ^{4}$He.   The fluid is relatively cold,
$T = 10^7$ K, has density $\rho = 5 \times 10^6$ g/cm$^3$, and is traveling toward the
origin with velocity $-2 \times 10^8$ m/s. A uniform constant gravitational acceleration
is applied, $g = -1.1 \times 10^8$ m/s$^{2}$. This setup causes a sharp initial release
of energy at $x = 0$, and the primary question is whether a detonation occurs promptly
near this contact point, or occurs later (possibly at a distance from the contact point).
The simulated domain has width $1.6384 \times 10^9$ cm, and we apply inflow boundary conditions
that keep feeding the domain with material that has the same conditions as the initial fluid.
Simulations are performed with the adaptive mesh refinement (AMR) code \castro.
For the burning we use the alpha-chain nuclear network
\texttt{aprox13}. \deleted{The results we describe below do not strongly depend on the network.}
Release 18.12 of the \castro\ code was used. The \amrex\ and \microphysics\ repositories
that \castro\ depends on were also on release 18.12. The problem is located in the
\texttt{Exec/science/Detonation} directory, and we used the \texttt{inputs-collision} setup.

The simulation is terminated when the peak temperature on the domain first reaches
$4 \times 10^9$ K, which we call a thermonuclear ignition (\added{for reference, the
density at the location where the ignition occurs is approximately $1.4\times 10^7\ \text{g / cm}^3$}). This stopping criterion is a
proxy for the beginning of a detonation. Reaching this temperature does not guarantee
that a detonation will begin, and in this study we do not directly address the question
of whether a ignition of this kind always leads to a detonation. \added{Nor are we commenting
on the physics of the ignition process itself.} Rather, the main question
we investigate here is whether this ignition is numerically converged, \added{and for this purpose
this arbitrary stopping point is sufficient, since in a converged simulation the stopping point
should be reached at the same time independent of resolution}. A converged ignition
is a prerequisite to having a converged detonation. We measure two diagnostic quantities:
the time since the beginning of the simulation required to reach this ignition criterion,
and the distance from the contact point of the peak temperature.

@@

\autoref{fig:self-heat-distance} shows our main results. The lowest resolution we consider,
256 km, is typical of the early simulations of white dwarf collisions, and demonstrates a
prompt ignition near the contact point. As the (uniform) resolution increases, the ignition
tends to occur earlier and nearer to the contact point. This trend is not physically meaningful:
all simulations with resolution worse than about 1 km represent the same prompt central ignition,
and as the resolution increases, there are grid points physically closer to the center that can ignite.
However, when the resolution is better than 1 km, the situation changes dramatically: the prompt
central ignition does not occur, but rather the ignition is delayed and occurs further from the contact
point. When we have finally \replaced{(perhaps) converged}{reached the point where the curves start to flatten
and perhaps begin to converge}, 
the ignition occurs around 900 km from the contact point, about 1 second after contact (contrast to less than
0.05 seconds for the simulation with 1 km resolution). \added{Even at this resolution, it is not clear if
the simulation is converged. We were unable to perform higher resolution simulations to check convergence
due to the length of time that would be required.} This story contains two important lessons. First,
the required resolution for even a qualitatively converged simulation, less than 100 m, is out of reach
of current 3D supernova simulations. Second, the behavior for resolutions worse than 1 km qualitatively
appears to be converged, and one could perhaps be misled into thinking that there was no reason to
try higher resolutions, which is reason for caution in interpreting reacting hydrodynamics simulations.

@@

\citet{kushnir:2013} observe an important possible failure mode
for reacting hydrodynamics simulations. Let us define $\tau_{\rm e} = e / \dot{e}$
as the nuclear energy injection timescale, and $\tau_{\rm s} = \Delta x / c_{\rm s}$
as the sound-crossing time in a zone (where $\Delta x$ is the grid
resolution and $c_{\rm s}$ is the speed of sound). When the sound-crossing
time is too long, energy is built up in a zone faster than it can be
advected away by pressure waves. \added{This effect generalizes to
Langrangian simulations as well, where $\tau_{\rm s}$ should be understood
as the timescale for transport of energy to a neighboring fluid element.}
This is of course a problem inherent 
only to numerically discretized systems as the underlying fluid equations
are continuous. This can lead to a numerically seeded detonation
caused by the temperature building up too quickly in the zone; the
detonation may be spurious in this case. If $\tau_{\rm s} \ll \tau_{\rm e}$,
we can be confident that a numerically seeded detonation has not
occurred. In practice, we will quantify this requirement as:
\begin{equation}
  \tau_{\rm s} \leq f_{\rm s}\, \tau_{\rm e} \label{eq:burning_limiter}
\end{equation}
and require that $f_{s}$ is sufficiently smaller than one.
\citet{kushnir:2013} state that $f_{\rm s} = 0.1$ is a sufficient
criterion for avoiding premature ignitions. \citeauthor{kushnir:2013}
enforced this criterion on their simulations by artificially limiting
the magnitude of the energy release after a burn, and claimed that
this is resulted in more accurate WD collision simulations.

We find that for our test problem (and also the WD collisions we have simulated)
we do observe $\tau_{\rm s} > \tau_{\rm e}$; typically the ratio is a factor of 2--5 at
low resolution (\added{see \autoref{fig:self-heat-ts_te}}). This means that an ignition is
very likely to occur for numerical reasons, regardless of whether it would occur for physical reasons.
\added{At low resolution, adding more resolution does not meaningfully improve the ratio of
$\tau_s$ to $\tau_e$ at the point of ignition. The ignition timescale is so short
that virtually all of the energy release occurs in a single timestep even if the
timestep gets shorter due to the CFL limiter. It is only when the resolution gets
sufficiently high that we can simultaneously resolve the energy release over multiple
timesteps and the advection of energy across multiple zones. Even at the highest resolution
we could achieve, about 50 cm, $\tau_s / \tau_e$ was 0.8 at ignition, which is not
sufficiently small to be confident of numerical stability at that resolution. Note
that merely decreasing the timestep (at fixed resolution) does not help here either, as the
instability criterion is, to first order, independent of the size of the timestep.}

@@

We thus investigate whether limiting the energy release of the burn (we will term this
``suppressing'' the burn), as proposed by \citeauthor{kushnir:2013},
is a useful technique for avoiding the prompt detonation. \added{Since the limiter ensures
the inequality in \autoref{eq:burning_limiter} holds by construction, the specific
question to ask is whether the limiter achieves the correct answer and is converged
in cases where the simulation would otherwise be uncorrect or unconverged.}

@@

Our example detonation problem demonstrates, at least for this class of
hydrodynamical burning problem, a grid resolution requirement much more stringent
than 1 km. \replaced{This does not represent all such burning problems: for example, we
have tested similar setups with only carbon and oxygen for which a resolution
in the tens of kilometers is sufficient.}{This test does not, of course, represent
all possible WD burning conditions.}
However, the fact that it is even possible for
\replaced{this type of burning}{burning in white dwarf material} to require a
resolution better than 100 m should cast significant uncertainty on the results
of simulations featuring detonations that do not meet this resolution,
\added{regardless of the specific conditions). Or, at least, it should suggest
that stronger demonstrations of convergence are required (especially bearing in mind
our observation that the numerical instability can result in simulations that
appear qualitatively converged when the resolution is increased by a factor of
one or two orders of magnitude but not three orders of magnitude.}

>>>>

Early studies on WD collisions \citep{rosswog:2009,raskin:2010,loren-aguilar:2010,
hawley:2012,garcia-senz:2013} typically had effective spatial resolutions in the
range 100--500 km for the grid codes, and 10--100 km for the SPH codes, and observed
detonations that convert a large amount of carbon/oxygen material into iron-group elements.
These studies varied in methodology (Lagrangian versus Eulerian evolution, nuclear network
used) and did not closely agree on the final result of the event (see Table 4 of
\cite{garcia-senz:2013} for a summary). \cite{kushnir:2013} argued that many of these
simulations featured numerically unstable evolution, ultimately caused by the zone size
being significantly larger than the length scale over which detonations form. The detonation
length scale can vary widely based on physical conditions \citep{seitenzahl:2009,garg:2017}
but is generally not larger than 10 km.

In this paper, we attempt to find what simulation length scale is required to achieve
converged thermonuclear ignitions. The inspiration for this work comes from our
simulations of WD collisions using the reactive hydrodynamics code \castro\
\citep{castro, astronum:2017}. We have done both 2D axisymmetric and 3D simulations
of collisions of $0.64\ \msolar$ carbon/oxygen WDs, and we were unable to achieved converged
simulations at any resolution we could afford to run (the best was an effective zone size of
0.25 km, using adaptive mesh refinement, for the 2D case). We were therefore forced to
turn to 1D simulations, where we can achieve much higher resolution (at the cost, of course,
of not being able to do a test that can be directly compared to multi-dimensional simulations).
Our goal is to record our thoughts on why we believe we did not achieve convergence.

@@

Our test problem is inspired by \cite{kushnir:2013}, and very loosely approximates the
conditions of two $0.64\ \msolar$ WDs colliding head-on. The simulation domain is 1D with a
reflecting boundary at $x = 0$. For $x > 0$ there is a uniform fluid composed\footnote{
We note that the WD simulations that inspired this were pure carbon/oxygen,
but for the setup we describe here, pure carbon/oxygen conditions do not detonate at all.
This is not particularly surprising, since the 1D setup is a very imperfect representation
of the real multi-dimensional case, and is missing multi-dimensional hydrodynamics that could
substantially alter the dynamical evolution. So we added a small amount of helium to ensure
that this setup ignites. Of course, there will likely be a small amount of helium present in
C/O white dwarfs as a remnant of the prior stellar evolution.} (by mass)
of $50\%\, ^{12}$C, $45\%\, ^{16}$O, and $5\%\, ^{4}$He.   The fluid is relatively cold,
$T = 10^7$ K, has density $\rho = 5 \times 10^6$ g/cm$^3$, and is traveling toward the
origin with velocity $-2 \times 10^8$ m/s. A uniform constant gravitational acceleration
is applied, $g = -1.1 \times 10^8$ m/s$^{2}$. This setup causes a sharp initial release
of energy at $x = 0$, and the primary question is whether a detonation occurs promptly
near this contact point, or occurs later (possibly at a distance from the contact point).
The simulated domain has width $1.6384 \times 10^9$ cm, and we apply inflow boundary conditions
that keep feeding the domain with material that has the same conditions as the initial fluid.
Simulations are performed with the adaptive mesh refinement (AMR) code \castro.
For the burning we use the alpha-chain nuclear network \texttt{aprox13}.
Release 18.12 of the \castro\ code was used. The \amrex\ and \microphysics\ repositories
that \castro\ depends on were also on release 18.12. The problem is located in the
\texttt{Exec/science/Detonation} directory, and we used the \texttt{inputs-collision} setup.

The simulation is terminated when the peak temperature on the domain first reaches
$4 \times 10^9$ K, which we call a thermonuclear ignition (for reference, the
density at the location where the ignition occurs is approximately $1.4\times 10^7\ \text{g / cm}^3$). This stopping criterion is a
proxy for the beginning of a detonation. Reaching this temperature does not guarantee
that a detonation will begin, and in this study we do not directly address the question
of whether a ignition of this kind always leads to a detonation. Nor are we commenting
on the physics of the ignition process itself. Rather, the main question
we investigate here is whether this ignition is numerically converged, and for this purpose
this arbitrary stopping point is sufficient, since in a converged simulation the stopping point
should be reached at the same time independent of resolution. A converged ignition
is a prerequisite to having a converged detonation. We measure two diagnostic quantities:
the time since the beginning of the simulation required to reach this ignition criterion,
and the distance from the contact point of the peak temperature.

@@

\autoref{fig:self-heat-distance} shows our main results. The lowest resolution we consider,
256 km, is typical of the early simulations of white dwarf collisions, and demonstrates a
prompt ignition near the contact point. As the (uniform) resolution increases, the ignition
tends to occur earlier and nearer to the contact point. This trend is not physically meaningful:
all simulations with resolution worse than about 1 km represent the same prompt central ignition,
and as the resolution increases, there are grid points physically closer to the center that can ignite.
However, when the resolution is better than 1 km, the situation changes dramatically: the prompt
central ignition does not occur, but rather the ignition is delayed and occurs further from the contact
point. When we have finally reached the point where the curves start to flatten and perhaps begin to converge, 
the ignition occurs around 900 km from the contact point, about 1 second after contact (contrast to less than
0.05 seconds for the simulation with 1 km resolution). Even at this resolution, it is not clear if
the simulation is converged. We were unable to perform higher resolution simulations to check convergence
due to the length of time that would be required. This story contains two important lessons. First,
the required resolution for even a qualitatively converged simulation, less than 100 m, is out of reach
of current 3D supernova simulations. Second, the behavior for resolutions worse than 1 km qualitatively
appears to be converged, and one could perhaps be misled into thinking that there was no reason to
try higher resolutions, which is reason for caution in interpreting reacting hydrodynamics simulations.

@@

\citet{kushnir:2013} observe an important possible failure mode
for reacting hydrodynamics simulations. Let us define $\tau_{\rm e} = e / \dot{e}$
as the nuclear energy injection timescale, and $\tau_{\rm s} = \Delta x / c_{\rm s}$
as the sound-crossing time in a zone (where $\Delta x$ is the grid
resolution and $c_{\rm s}$ is the speed of sound). When the sound-crossing
time is too long, energy is built up in a zone faster than it can be
advected away by pressure waves. This effect generalizes to
Langrangian simulations as well, where $\tau_{\rm s}$ should be understood
as the timescale for transport of energy to a neighboring fluid element.
This is of course a problem inherent 
only to numerically discretized systems as the underlying fluid equations
are continuous. This can lead to a numerically seeded detonation
caused by the temperature building up too quickly in the zone; the
detonation may be spurious in this case. If $\tau_{\rm s} \ll \tau_{\rm e}$,
we can be confident that a numerically seeded detonation has not
occurred. In practice, we will quantify this requirement as:
\begin{equation}
  \tau_{\rm s} \leq f_{\rm s}\, \tau_{\rm e} \label{eq:burning_limiter}
\end{equation}
and require that $f_{s}$ is sufficiently smaller than one.
\citet{kushnir:2013} state that $f_{\rm s} = 0.1$ is a sufficient
criterion for avoiding premature ignitions. \citeauthor{kushnir:2013}
enforced this criterion on their simulations by artificially limiting
the magnitude of the energy release after a burn, and claimed that
this is resulted in more accurate WD collision simulations.

We find that for our test problem (and also the WD collisions we have simulated)
we do observe $\tau_{\rm s} > \tau_{\rm e}$; typically the ratio is a factor of 2--5 at
low resolution (see \autoref{fig:self-heat-ts_te}). This means that an ignition is
very likely to occur for numerical reasons, regardless of whether it would occur for physical reasons.
At low resolution, adding more resolution does not meaningfully improve the ratio of
$\tau_s$ to $\tau_e$ at the point of ignition. The ignition timescale is so short
that virtually all of the energy release occurs in a single timestep even if the
timestep gets shorter due to the CFL limiter. It is only when the resolution gets
sufficiently high that we can simultaneously resolve the energy release over multiple
timesteps and the advection of energy across multiple zones. Even at the highest resolution
we could achieve, about 50 cm, $\tau_s / \tau_e$ was 0.8 at ignition, which is not
sufficiently small to be confident of numerical stability at that resolution. Note
that merely decreasing the timestep (at fixed resolution) does not help here either, as the
instability criterion is, to first order, independent of the size of the timestep.

@@

We thus investigate whether limiting the energy release of the burn (we will term this
``suppressing'' the burn), as proposed by \citeauthor{kushnir:2013},
is a useful technique for avoiding the prompt detonation. Since the limiter ensures
the inequality in \autoref{eq:burning_limiter} holds by construction, the specific
question to ask is whether the limiter achieves the correct answer and is converged
in cases where the simulation would otherwise be uncorrect or unconverged.

@@

Our example detonation problem demonstrates, at least for this class of
hydrodynamical burning problem, a grid resolution requirement much more stringent
than 1 km. This test does not, of course, represent all possible WD burning conditions.
However, the fact that it is even possible for burning in white dwarf material to require a
resolution better than 100 m should cast significant uncertainty on the results
of simulations featuring detonations that do not meet this resolution,
regardless of the specific conditions). Or, at least, it should suggest
that stronger demonstrations of convergence are required (especially bearing in mind
our observation that the numerical instability can result in simulations that
appear qualitatively converged when the resolution is increased by a factor of
one or two orders of magnitude but not three orders of magnitude.
